\Large{All in the family:\\German twin finishing times in\\the 2016 women's Olympic marathon}

### 2017-02-23 paper/chance_memo.tex
misc edits

\documentclass[12pt]{article}
\sloppy
\usepackage{times}
\usepackage{url}
\usepackage{arydshln}

@@

@@

@@

@@

  We changed the language in the paper so it uses this phrasing.

@@

\item There are errors in figure labels, ``5a'' and ``5b''

@@

  We have changed the Latin to, all things equal.

@@

>>>>

\documentclass[12pt]{article}
\sloppy
\usepackage{times}
\usepackage{xr-hyper}
\usepackage{url}
\usepackage{arydshln}

\externaldocument{paper}

@@

  The $p$-value for the quadratic term is -1.88, which is not
  particularly compelling.  We have accordingly dropped the quadratic
  terms in the scatter plots on p.\ 4 and, implicitly, from the
  Studentized residual plots on p.\ 6.
  %% Note: xr labels are
  %% breaking; looks like some sort of
  %% label clash with another package.
  %% Hardcoding here is poor style.

  % > fit <- lm (FINAL ~ PB + I(PB^2), data = df)
  % > print (summary(fit))

  % Call:
  % lm(formula = FINAL ~ PB + I(PB^2), data = df)

  % Residuals:
  % Min      1Q  Median      3Q     Max 
  % -817.64 -288.70  -92.13  224.05 1319.33

  % Coefficients:
  % Estimate Std. Error t value Pr(>|t|)   
  % (Intercept) -1.257e+04  6.417e+03  -1.958  0.05233 . 
  % PB           3.633e+00  1.336e+00   2.719  0.00744 **
  % I(PB^2)     -1.309e-04  6.946e-05  -1.884  0.06181 . 
  % ---
  % Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

  % Residual standard error: 417.5 on 130 degrees of freedom
  % (24 observations deleted due to missingness)
  % Multiple R-squared:  0.6538,	Adjusted R-squared:  0.6484 
  % F-statistic: 122.7 on 2 and 130 DF,  p-value: < 2.2e-16

@@

  We have added age to our simulation model, and the results are
  qualitatively equivalent.  In addition, if one adds age to the
  linear specification depicted in our basic scatter plot (now without
  a quadratic term, as noted above), age is a significant predictor of
  marathon finishing time, even controlling for personal best
  ($p \approx 2.426$).  Older runners are slower, \emph{ceteris
    paribus}.

% Call:
% lm(formula = FINAL ~ PB + years2marathon, data = df)

% Residuals:
%     Min      1Q  Median      3Q     Max 
% -779.88 -281.31  -94.71  222.21 1319.29

% Coefficients:
%                  Estimate Std. Error t value Pr(>|t|)    
% (Intercept)    -1.366e+03  7.430e+02  -1.838   0.0684 .  
% PB              1.147e+00  7.222e-02  15.876   <2e-16 ***
% years2marathon  1.799e+01  7.413e+00   2.426   0.0166 *  
% ---
% Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

% Residual standard error: 413.9 on 130 degrees of freedom
%   (23 observations deleted due to missingness)
% Multiple R-squared:  0.6597,	Adjusted R-squared:  0.6545 
% F-statistic:   126 on 2 and 130 DF,  p-value: < 2.2e-16

@@

  We have changed the color scheme for the dots in our figures.  The
  German Hahner twins are now shown in black, the North Korean Kim
  twins in red, and the Estonian Luik triplets in blue.  Each color is
  present in its country's respective flag.

@@

  We changed the language in the paper so that it uses this phrasing.

@@

  We fully agree with this comment, and indeed the intuition offered
  by the referee was the motivation behind our simulation.  To make
  this clear, we added a paragraph after the simulation steps, and
  this paragraph explains why finishing times (and order) will vary
  both within and across simulations.  Then, the plots we generated
  (see Figures 5 and 6) describe how often the Hahner twins finish
  in close proximity of each other.

  We have thought about historical marathon data but are hesitant to
  go down that route.  The set of runners who competed in Rio is
  unique to this race, and marathon finishes are confounded by course
  and local conditions.

\item There are errors in figure labels ``5a'' and ``5b''

@@

  We have changed the Latin text to, all things equal.

@@

  We apologize for the confusion here, but we are not sure what the
  referee is requesting.  In the marked-up PDF document, the word
  ``likely'' appears over the following sentence: ``Their finish—one
  second between the two women—was an extremely low probability
  event.''  We are not sure how to incorporate the referee's request
  but are happy to do so if given more information.

### 2017-02-23 paper/paper.tex
misc edits

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain second-order polynomial regression lines.  The ticks along the
horizontal axes in both plots indicate times, either personal bests or
half marathon splits, of runners who earned DNF results.

\begin{figure}[!ht]
  \centering
  \caption{Olympic finishing times as a function of underlying marathon talent}
  \label{fig:scatter}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot.pdf}
    \caption{Based on personal best} 
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half.pdf}
    \caption{Based on half marathon split} 
    \label{fig:45degreeplot_half}
  \end{subfigure}
\end{figure}

@@

Another perspective on the extent to which the Hahners' overall
marathon finishing times were not unusual can be gleaned from Figure
\ref{fig:residuals}.  This figure presents cumulative distributions of
the polynomial regression residuals from the regression models
displayed in the two panels of Figure \ref{fig:scatter}.  In
particular, we calculated the residuals from Figures
\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
them, and then arranged resulting Studentized residuals from least to
most along the horizontal axes in Figures
\ref{fig:studentizedresiduals} and \ref{fig:studentizedresiduals_half}.

\begin{figure}[!ht]
  \centering
  \caption{Studentized residuals from Olympic finishing time regressions}
  \label{fig:residuals}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals.pdf}
    \caption{Based on personal best} 
    \label{fig:studentizedresiduals}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half.pdf}
    \caption{Based on half marathon split}
    \label{fig:studentizedresiduals_half}
  \end{subfigure}
\end{figure}

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's red dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to subsequent DNFs.  The usual color scheme applies here,
and the Estonian triplet DNF prior to the half marathon split is
evident. As noted in the figure, Lisa Hahner was ahead of her sister
through 15 kilometers at which point Anna took the lead.  Figure
\ref{fig:secondsbehind} contains two orange dots representing the North
Korean Kim twins, but this is not visually apparent because the Kim
twins had identical split times during the entire marathon.

@@

Our visualizations are useful and shed a fair bit of light on the
Hahner twin's performance in the 2016 women's Olympic marathon.  In
the interest of increasing precision, we now pose the following
question: if we take into consideration the Hahner twin's similarities
in marathon talent as well as natural variation in marathon finishing
times, what is the probability that the twins would finish the Rio
race at roughly the same time and/or sequentially?  To answer this
question, we need to know the counterfactual distribution of potential
marathon finishing times that would have occurred had the Hahner twins
independently (in particular, of each other) and repeatedly run the
Rio marathon, holding constant marathon conditions, the abilities of
other runners, and so forth.  Access to such a distribution would
establish the set of potential outcomes that could have occurred on
August 14, 2016, and we could in principle use this distribution to
determine the likelihood that a simultaneous finish by the Hahners, or
at least a near-simultaneous finish, occurred by chance alone.  If
these twins rarely finish the marathon together in such a
counterfactual world, then one might be skeptical that their observed
finish occurred without coordination.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
to establish a distribution of potential race outcomes for the Hahner
twins.  However, we can attempt to simulate this distribution by
estimating the distribution of every other runner's finishing time,
conditional on marathon talent, and then drawing from this
distribution in order to calculate the likelihood that, for example,
Lisa and Anna Hahner finished simultaneously.

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of the runner's marathon skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal distribution. Taking each runner's 
personal best marathon time as a proxy for their marathon skill, we
estimate $\beta_1$, $\beta_2$, $\beta_3$ and $\sigma$ using ordinary least
squares on all finishing Rio marathoners.  We exclude the Hahner/Kim
twins and Luik triplets from the sample so that that our estimates are not affected by any intentional twin/triplet finish.  
For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$, her best time squared $X^2_{i}$, and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and $\tilde{\beta}_3$,
  respectively, from a multivariate normal distribution with mean
  equal to the previously estimated coefficient vector and covariance
  equal to the previously estimated covariance matrix.
\item Given values of $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
  $\tilde{\beta}_3$, for each runner draw an error $\tilde{e}$ from a
  normal distribution with mean zero and a standard deviation equal to
  the standard deviation of the original regression model's residuals.
\item For each runner, predict her final result by combining the
  randomly generated beta coefficients and error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{\beta}_3\,Z_i +  \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

>>>>

Figure \ref{fig:scatter} contains two plots, both of which describe
how finishing times from the Rio women's marathon varied as a function
of athletes' personal best times and half marathon split times.  The
points in the plots are colored by twin/triplet status, and both plots
contain second-order polynomial regression lines.  The ticks along the
horizontal axes in both plots indicate times, either personal bests or
half marathon splits, of runners who earned DNF
results.\label{scatterplots}

\begin{figure}[!ht]
  \centering
  \caption{Olympic finishing times as a function of underlying marathon talent}
  \label{fig:scatter}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_no_quadratic.pdf}
    \caption{Based on personal best} 
    \label{fig:45degreeplot}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{scatter_plot_half_no_quadratic.pdf}
    \caption{Based on half marathon split} 
    \label{fig:45degreeplot_half}
  \end{subfigure}
\end{figure}

@@

Another perspective on the extent to which the Hahners' overall
marathon finishing times were not unusual can be gleaned from Figure
\ref{fig:residuals}.  This figure presents cumulative distributions of
the polynomial regression residuals from the regression models
displayed in the two panels of Figure \ref{fig:scatter}.  In
particular, we calculated the residuals from Figures
\ref{fig:45degreeplot} and \ref{fig:45degreeplot_half}, Studentized
them, and then arranged resulting Studentized residuals from least to
most along the horizontal axes in Figures
\ref{fig:studentizedresiduals} and
\ref{fig:studentizedresiduals_half}.\label{residualplots}

\begin{figure}[!ht]
  \centering
  \caption{Studentized residuals from Olympic finishing time regressions}
  \label{fig:residuals}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_no_quadratic.pdf}
    \caption{Based on personal best} 
    \label{fig:studentizedresiduals}
  \end{subfigure}
  \begin{subfigure}{.495\textwidth}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{studentized_residuals_half_no_quadratic.pdf}
    \caption{Based on half marathon split}
    \label{fig:studentizedresiduals_half}
  \end{subfigure}
\end{figure}

@@

We can generalize this result by looking at all pairs of runners in
the marathon.  For all 8,778 pairs of 133 finishers, Figure
\ref{fig:diffdiffscatter} plots differences in finishing times against
differences in personal best times, and pairs of twins/triplets are
identified by the same color scheme we used earlier.  Consider first
the Hahner twins.  The two German women are effectively located on the
horizontal axis because their difference in finishing times is one
second.  However, there are many points above the Hahner's black dot,
and this shows that, conditional on an approximate 115 second
difference in personal best times, most marathoners did not have close
finishing times like the Hahners.  Some pairs of runners with around
115 second personal best differences had finishing time differences of
1,000 seconds, i.e., in excess of 15 minutes.  Of course the points in
Figure \ref{fig:diffdiffscatter} are not independent, but they provide
a sense of the dispersion in finishing time differences that one can
expect conditional on differences in personal best times.

@@

Figure \ref{fig:secondsbehind} provides a sense of the lockstep nature
of the Kim twins, and it describes each Olympic runner's status at
various split times on the marathon course.  Each dot in the figure
depicts a recorded split and the number of seconds each runner was
behind the race leader at the time.  There are more dots at earlier
splits due to subsequent DNFs.  The usual color scheme applies here,
and the Estonian triplet DNF prior to the half marathon split is
evident. As noted in the figure, Lisa Hahner was ahead of her sister
through 15 kilometers at which point Anna took the lead.  Figure
\ref{fig:secondsbehind} contains two red dots representing the North
Korean Kim twins, but this is not visually apparent because the Kim
twins had identical split times during the entire marathon.

@@

Our visualizations shed a fair bit of light on the Hahner twin's
performance in the 2016 women's Olympic marathon.  In the interest of
increasing precision, we now pose the following question: if we take
into consideration the twin's similarities in marathon talent as well
as natural variation in marathon finishing times, what is the
probability that they would finish the Rio race at roughly the same
time and/or sequentially?  To answer this question, we need to know
the counterfactual distribution of potential marathon finishing times
that would have occurred had the Hahner twins independently (in
particular, of each other) and repeatedly run the Rio marathon,
holding constant marathon conditions, the abilities of other runners,
and so forth.  Access to such a distribution would establish the set
of potential outcomes that could have occurred on August 14, 2016, and
we could in principle use this distribution to determine the
likelihood that a simultaneous finish by the Hahners, or at least a
near-simultaneous finish, occurred by chance alone.  If these twins
rarely finish the marathon together in such a counterfactual world,
then one might be skeptical that their observed finish occurred
without some degree of coordination.

Unfortunately for us but nonetheless fortunately for the race's
participants, it is not possible to rerun the women's Olympic marathon
to establish a distribution of potential race outcomes for the Hahner
twins.  However, we can attempt to simulate this distribution by
estimating the distribution of every other runner's finishing time,
conditional on marathon talent, and then drawing from this
distribution in order to calculate the likelihood that, for example,
Lisa and Anna Hahner finished the Rio race simultaneously.

To estimate the conditional distribution of each Rio final result, we
assume that each runner's marathon time \Yi\ is distributed normally
with a mean that is a quadratic function of the runner's marathon
skill $X_i$ and her age $Z_i$. Hence:
$$\Yi \mid X_i \sim N\left(\beta_0 + \beta_1\,X_{i} + \beta_2\,X^2_{i}  + \beta_3\,Z_{i} + e_i, \sigma\right),$$
%%
where $N\left(\cdot,\cdot\right)$ denotes a normal
distribution. Taking each runner's personal best marathon time as a
proxy for her underlying marathon skill, we estimate $\beta_1$,
$\beta_2$, $\beta_3$ and $\sigma$ using ordinary least squares on all
finishing Rio marathoners.  We exclude the Hahner/Kim twins and Luik
triplets from the sample so that that our estimates are not affected
by the twin/triplet finishes which, theoretically, could reflect the
results of coordination.  For a simulated marathon, we draw a runner's
time from an estimated distribution and condition on the runner's
personal best marathon time and age.  Once a race is simulated for all
runners, twins and triplets included, we record both the time between
Anna and Lisa Hahner's simulated finishes and the difference in their
simulated ranks.  We then simulate a new race---drawing a new set of
finishing times---and record the same quantities. The steps of the
simulation are as follows.
\begin{enumerate}
\item Ignoring twins and triplets, estimate with least squares a
  linear model that predicts a runner's finishing time $Y_i$ based on
  her pre-Olympic personal best time $X_i$, her best time squared
  $X^2_{i}$, and her age $Z_i$.
\item Extract the resulting coefficient vector, estimated covariance
  matrix, and estimated regression variance from this model.
\item For each simulated rate, draw intercept and slope estimates
  $\tilde{\beta}_0$, $\tilde{\beta}_1$, $\tilde{\beta}_2$, and
  $\tilde{\beta}_3$, respectively, from a multivariate normal
  distribution with mean equal to the previously estimated coefficient
  vector and covariance equal to the previously estimated covariance
  matrix.
\item For each runner, draw an error $\tilde{e}$ from a normal
  distribution with mean zero and a standard deviation equal to the
  standard deviation of the original regression model's residuals.
\item Predict each runner's final result by combining the randomly
  generated beta coefficients and individual error terms,
  $\tilde{y} = \tilde{\beta}_0 + \tilde{\beta}_1\,X_i +
  \tilde{\beta}_2\,X_i^2 + \tilde{\beta}_3\,Z_i + \tilde{e}$.
\item Eliminate each runner from the simulated race with a probability
  equal to the observed fraction of marathoners who did not finish the
  marathon.
% \item Calculate the difference in time and difference in ranking
%   between Anna Hahner and Lisa Hahner, assuming both finished the
%   race.
\item Repeat above steps 10,000 times.
\end{enumerate}

As these steps illustrate, our simulation repeatedly draws random
coefficient vectors, and this captures variability in what we know
about the relationship between runner talent and age and runer finish
times.  In addition, the simulation draws random disturbances for each
runner, conditional on our original estimate of regression variance;
these disturbances capture variability in runner finishing times,
notwithstanding underlying runner talent.  Importantly, the
disturbances that we draw are independent across runners.  This means
that the finishing order among runners for each simulated race will
vary.
